Towards typed repositories of proofs \\[0.6em] \small \textsf{MIPS 2010}

### 2010-07-09 talks/MIPS2010/slides.tex
exemple gasp, qqes détails
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@147 403e655e-980a-47bd-8c60-1ee5348bb6b7
Version finale des slides (?)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@148 403e655e-980a-47bd-8c60-1ee5348bb6b7

% Mode d'emploi: Switcher les deux lignes suivantes pour passer du
% mode présentation au handout
% ===================================================
\documentclass[ignorenonframetext,red]{beamer}
% \documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
% ===================================================

@@

This talk will be about some remarks and directions on the way the
edition of formal proof is done in proof assistants.

@@

The paradox about this situation is that these very tools can be used
to reason about formal languages and especially proof languages: some
of these tools are even dedicated to the mechanization of the
metatheory of proof languages. But we still use the legacy generic
tools to manage our developments.

@@

\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \item Dependency
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, express the incrementality of
the constructions, and their dependency. It can be viewed as a kernel
for a generic, typed version control system.

\begin{frame}{Inspiration: the \textsf{git} storage model}
  \begin{center}
    \begin{overlayarea}{\textwidth}{12em}
      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
      \only<10>{
        \begin{center}
          \vspace{5em}
          {\Large Let's design a \emph{typed} \textsf{git}}
        \end{center}
      }%
    \end{overlayarea}
  \end{center}%
  \begin{itemize}\small %
  \item<7-> ``Content-adressable''%
  \item<8-> Name reflects content%
  \item<9-> Maximal sharing (or hash-consing)%
  \end{itemize}%
\end{frame}

\textsf{Git} is one of the most successful version control system, and
part of its success is due to the very simple, yet very powerful
storage model. It doesn't keep track of changes \emph{per se}, but of
similarities between version. Given a base directory structure,
already committed in its database, and a new version of it with, let's
say, one file changed -- \texttt{f3.txt}, it stores the new version by
linking all common subdirectory structure to the previous version.
This is possible because every node in the tree, directories and
files, have a unique name. So actually, the database of \textsf{git}
looks like this: it associates to every name, a content (for files) or
a list of names (for directories). The invariant is that this database
forms a DAG. This naming policy makes all content adressable.

@@

\begin{frame}{A typed repository}
  % exemple
\end{frame}

on va plus loin grace a l'approche LF où tous les objets ont un type,
et font partie de l'arbre de syntaxe. Donc tout est hash-consé
(règles logiques, syntaxe, preuves).

\begin{frame}{Incremental type-checking, incrementally}%
  \begin{block}{Syntax}%
    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
      |\ *\] \[\]}%
    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
      \alert{t\ t}\ |\ *\] \[\]}%
    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
      *\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
      \alert{(x = a)\cdot t}\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
      |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
  \end{block}
  \begin{block}{Semantics}
      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
        \alert{\Gamma[x=a:t]}}
      \only<5->{\ |\ \Gamma[x=a:t]}\] 
      \vspace{1em}
      \begin{center}
        \begin{overlayarea}{0.8\textwidth}{5em}
          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
            the new repository $\Delta$}'' }
        \end{overlayarea}
      \end{center}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Product - Init}
  \begin{mathpar}
    \large
    \infer                      % prod
    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
      *\uncover<2->{\Rightarrow\Delta}}
    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
    \and
    \uncover<3->{                               %init
      \infer{ }
      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
      \quad {\small [x:t]\in\Gamma}
    }
  \end{mathpar}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Equality binder}
    \begin{mathpar}
      \large
      \infer
      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
        *\Rightarrow\Delta }
      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \notin\Gamma}
      \and \pause
      \infer
      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \in\Gamma}
    \end{mathpar}
  \end{block}
  \pause
  \begin{block}{Application}
    \[ \infer
      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
      \quad [x : u] \in \Gamma
      \]
  \end{block}
\end{frame}

\begin{frame}{Naming}

\end{frame}

environnement : nom -> contenu * type

\begin{frame}{Conclusion}

\end{frame}

>>>>

% Mode d'emploi: Switcher les deux lignes suivantes pour passer du
% mode présentation au handout
% ===================================================
\documentclass[ignorenonframetext,red]{beamer}
%\documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
% ===================================================

@@

I am going to present a work-in-progress, realized together with Yann
Régis-Gianas. It is far from a finished product, so feel free to
interrupt, comment or criticize. So this talk will be about some
observations and directions to improve the way we edit and store
formal proofs in proof assistants.

@@

The paradox about this situation is that these very tools (Coq,
Matita, Agda) can be used to reason about formal languages and
especially proof languages, some of these tools are even dedicated to
the mechanization of the metatheory of proof languages. But we still
use this legacy generic tools to manage our developments.

@@

\hrule
\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \item Dependency
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}
\hrule

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, express the incrementality of
the constructions, and their dependency. It can be viewed as a kernel
for a generic, typed version control system, and for the first
iteration of this project, we were inspired by the way \textsf{git}
stores the history of directories.

\hrule
\begin{frame}{Inspiration: the \textsf{git} storage model}
  \begin{center}
    \begin{overlayarea}{\textwidth}{12em}
      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
      \only<10>{
        \begin{center}
          \vspace{5em}
          {\Large Let's design a \emph{typed} \textsf{git}}
        \end{center}
      }%
    \end{overlayarea}
  \end{center}%
  \begin{itemize}\small %
  \item<7-> ``Content-adressable''%
  \item<8-> Name reflects content%
  \item<9-> Maximal sharing (or hash-consing)%
  \end{itemize}%
\end{frame}
\hrule

\textsf{Git} is one of the most successful version control system, and
part of its success is due to the very simple, yet very powerful
storage model. It doesn't keep track of changes \emph{per se}, but of
similarities between version. Given a base directory structure,
already committed in its database, and a new version of it with, let's
say, one file changed -- \texttt{f3.txt}, it stores the new version by
linking all common subdirectory structure to the previous
version. This way, similarities are physically relalized by this kind
of ``compression''. This is possible because every node in the tree,
directories and files, have a unique name. So actually, the database
of \textsf{git} looks like this: it associates to every name, a
content (for files) or a list of names (for directories). The
invariant is that this database forms a DAG. This naming policy makes
all content adressable.

@@

Let's apply the same discipline to proofs, and validate them by
typing. What would we get?

\hrule
\begin{frame}{A typed repository}
  \only<1>{\includegraphics[width=\textwidth]{images/gasp1.pdf}}%
  \only<2>{\includegraphics[width=\textwidth]{images/gasp2.pdf}}%
  \only<3>{\includegraphics[width=\textwidth]{images/gasp3.pdf}}%
  \only<4>{\includegraphics[width=\textwidth]{images/gasp4.pdf}}%
  \only<5>{\includegraphics[width=\textwidth]{images/gasp5.pdf}}%
\end{frame}
\hrule

On the same model, let's start with a proof of $C$ already committed,
and starting with this structure. $\pi_1$, $\pi_2$ and $\pi_3$ are the
rest of the proof, upwards. Each node is labelled with the logical
rule (or constructor) used, but also annotated by the judgement it
proves. Now let's imagine that we come up with a new, better proof for
$B$. Integrating it in the repository means, like before, to share all
common sub-derivations. But now, we don't need to retype the whole new
proof, just the new parts (in red). Like before, this is possible if
we give a name to all sub-derivations, and the database looks like
this: it associates with every name, how it is formed.

\hrule
\begin{frame}{Incremental type-checking, incrementally}%
  \begin{block}{Syntax}%
    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
      |\ *\] \[\]}%
    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
      \alert{t\ t}\ |\ *\] \[\]}%
    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
      *\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
      \alert{(x = a)\cdot t}\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
      |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
  \end{block}
  \begin{block}{Environments}
      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
        \alert{\Gamma[x=a:t]}}
      \only<5->{\ |\ \Gamma[x=a:t]} \] 
    \end{block}
    \begin{block}{Judgement}
      \vspace{1em}
        \begin{tabular}{ll} \Large
          $\Gamma\vdash t : u \uncover<7>{\alert{\Rightarrow\Delta}}$
          &
          \only<-6>{`` In environment $\Gamma$, term $t$ has type $u$
            ''}
          \uncover<7>{`` From
            \alert{repository} $\Gamma$, term $t$ of type $u$ \\ &
            \alert{\hspace{1.2em}leads to the new repository $\Delta$}''
          }
        \end{tabular}
    \end{block}
\end{frame}
\hrule

Let's construct a calculus for this. We start from the usual Pure Type
Systems, with abstraction, products, variables, applications and just
one sort for the sake of simplicity. Typing is done in the usual
name-type environment, by the usual judgement: ``In $\Gamma$, $t$ has
type $u$. We modify it gradually, keeping its expressivity but
tailoring the syntax to name all subterms.

First we don't want to allow compound, deep applications. We remove
general application and make it a list of variables. Then, we need a
new binder to remember these flat applications and give a name to
them. It will behave like a \textsf{let} binding, and we add a
corresponding entry to the environment. Note that in terms, the type
is not present whereas it is in the environment.

It happens that the system is already interesting enough for our
purpose without abstraction: the notions of dependency and
incrementality do not need the abstraction. For the moment, we remove
it, and with it all notion of computation.

These simple modifications allow us to see this system as a typed
repository of proofs with a last modification: the usual judgement
becomes ``From repository $\Gamma$, term $t$ of type $u$ leads to
repository $\Delta$''. The analogy here is: $\Gamma$ is the current
repository, $t$ is a piece of proof that has to be checked and
committed, it can have redundancies (it's not maximally shared), and
integration of $t$ into $\Gamma$ leads to $\Delta$, which is maximally
shared.

We thread the environment along the typing process and return it as
the result.

\newcommand\g[1]{\textcolor{gray}{#1}}

\hrule
\begin{frame}{Typing rules}
  \begin{block}{Product}
    \[ \large
    \infer                      % prod
    {\Gamma\vdash t : *\uncover<2->{ \textcolor{gray}{\ \Rightarrow \_}} \and \Gamma[x:t]\vdash u :
      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
    {\Gamma\vdash(x:t)\cdot u : * \uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
    \]
  \end{block}
  \pause\pause
  \begin{block}{Init}
    \[\infer{ }
      {\Gamma\vdash x : t \uncover<4->{\textcolor{gray}{\ \Rightarrow\Gamma}}}
      \quad {\small [x:t]\in\Gamma}
      \]
  \end{block}
\end{frame}
\hrule

I will present briefly some of the typing rules of this calculus. Here
is for example the usual product: we thread the return environment in
its main branch. Here is the usual init rule, we return the enlarged
environment.

\hrule
\begin{frame}{Typing rules}
  \begin{block}{Equality binder}
    \begin{mathpar}
      \large
      \infer
      {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} \and \Gamma[x=a:u]\vdash t :
        *\textcolor{gray}{\ \Rightarrow\Delta} }
      {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
      \quad {\small [y=a:u] \notin\Gamma}
      \and \pause
      \infer
      {\Gamma\vdash t \{y/x\} : * \textcolor{gray}{\ \Rightarrow\Delta}}
      {\Gamma\vdash(x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
      \quad {\small [y=a:u] \in\Gamma}
    \end{mathpar}
  \end{block}
  \pause
  \begin{block}{Application}
    \[ \infer
      {\Gamma\vdash a : (y:u)\cdot t \textcolor{gray}{\ \Rightarrow\Delta}}
      {\Gamma\vdash a\ x : t \{x/y\} \textcolor{gray}{\ \Rightarrow\Delta}}
      \quad [x : u] \in \Gamma
      \]
  \end{block}
\end{frame}
\hrule

The rules for the equality binder are two. The first resembles a lot
that of the \textsf{let} binding: we check the body, then the rest by
putting a new definition in the environment. But it applies only when
there is \emph{no} other equal definitions with body $a$ in the
environment. If there is, then the second rule apply.

In this case, we just ignore the first premise, as $a$ was already
type-checked. This rule is the key to incrementality.

Last rule, the application, is not very different from the usual one,
except that since there can be only a variable in the argument
position, we inline the Init rule and get rid of the second premise.

\hrule
\begin{frame}{Content-aware names (implementation)}
  Given ``$a$'', how to decide efficiently ``$[y = a : u] \in
  \Gamma$''?
  \pause
  \[ \begin{array}{lllr}
    | \cdot | &:& \vec\kappa \rightarrow \kappa & \text{Hash function}\\
    \equiv    &:& \kappa \rightarrow \kappa \rightarrow \mathbb{B} & \text{Efficient comparison}\\
    \nu       &:& \textsf{unit} \rightarrow \kappa & \text{Fresh key generator}
  \end{array} \]
  \pause
  \begin{mathpar}
    \infer
    {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} 
      \and \Gamma[|a|=a:u]\vdash t\{x/|a|\} :
      *\textcolor{gray}{\ \Rightarrow\Delta} }
    {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
    \quad {\small [y=a:u] \notin\Gamma}
    \and
    \pause
    \infer
    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[k:t]\vdash u\{x/k\} :
      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
    {\Gamma\vdash(x:t)\cdot u : * \textcolor{gray}{\textcolor{gray}{\ \Rightarrow\Delta}}}
    \quad{\small k=\nu()}
  \end{mathpar}
  \pause
  \Large
  \[ \Gamma\ :\ \kappa \to \vec\kappa * \tau \]
\end{frame}
\hrule

Let's delve a moment into a particular implementation detail: how to
choose names. We saw a quite unusual query to the environment before:
look for all variables with body exactly $a$. How to decide this query
efficiently?

We inspire once more from \textsf{git}, and give \emph{internal}
names, or keys, to all objects, that reflect their contents. We
introduce a morally injective hash function, from lists of keys to
keys, an efficient comparison function between two hashes, and a fresh
key generator.

Here are the introduction rules of the system: When we introduce a
definition $a$, we choose the hash of $a$ as a name for it, and do the
substitution in the rest of the term. When we introduce a declaration
(a product), we choose a new key for its name, and do the
substitution.

Therefore, the environment is then a function from a key to the
content of its definition (if it's a definition) and its type.

But let me finish with a remark: all of this is very static, content
is determined exactly by its syntactic shape. There seems to lack a
notion of computation into the calculus.

\hrule
\begin{frame}{Further Work}
  \Large
What if we reintroduce {\Huge $[x:t]\cdot t$} ?
\begin{enumerate} \normalsize
\pause
\item Constructive metatheory
\pause
\item A language to express patches?
\end{enumerate}
\end{frame}
\hrule

What would happen if we reintroduce computation, and in particular the
abstraction? We could not only declare logics but also develop its
metatheory, as we do in Pure Type Systems. For example, the weakening
property would have an existence, and we could even compute it since
it's constructively defined. In a sense, there is hope to relax the
very strict notion of equality implicit in the calculus, from
syntactic equality to ``semantic'' equality.

Moreover, we would be able to express functions that transform the
syntax, that is \emph{patches}. This opens a set of research
directions : are they valid with respect to a logic? what can we say
about dependencies between patches? Their commutation etc.
